A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-26 latex/main.tex
Updates from Overleaf

[TODO narrative is muddled here: the point of the para is to say that 
pango X lineage designation is manual and subjective.]
% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis. 
The first recombinant lineage XA was designated in 2021 based on the 
analysis of Jackson et al\.~\cite{Jackson2021}, which established 
that recombination with onward transmission had occured [weak FIXME].
As of pango-designation version v1.33.1 (2025-03-31), a total of 146 distinct 
X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146
Although the process of incorporating recombinant lineages in the Pango 
nomenclature system is well defined,
the identification of such events is much more subjective, requiring
the synthesis of multiple lines of evidence and often the 
manual inspection of phylogenies and sequence alignments
by volunteers.

By automatically detecting and incorporating recombination events
into the ARG, sc2ts allows us to model the origins of these 
X lineages as nodes in the graph. 
Table~\ref{tab:pango_x_lineages} summarises
these results for the XX Pango X lineages present in
the ARG. 
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
focal pango, many of which are the generic BA.2 label
[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but that do not meet the strict origination criteria for Type I.
Type III events are then nodes in the ARG associated with the 
orin

Of these, XAF and XJ are relatively straightforwardo

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All these deletions are known to occur major VOCs
(https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563;
\cite{Akaishi2023,Li2023indels}).
Eight of these deletions occur in S1-NTD,
a known hotspot for recurrent deletions \cite{McCarthy2021}.
There are three contiguous deletions in NSP6
(4-base at 11288, 5-base at 11283, and 5-base at 11292),
which should have been mapped as two 9-base deletions,
one above the BA.1 origin node and
the other above the BA.2 origin node in the ARG
(Figure~\ref{fig:subgraph_ba1_ba2}),
but they were broken up
during deletion remapping because they had
overlapping four sites at 11288-11292
(which was mapped as the 4-base deletion event).
Furthermore, the placement of these deletion events in the ARG
reflects that of the VOC-associated deletion events
inferred in a previous study \cite{Akaishi2023}.

We also found highly recurrent deletions using the ARG.
The top 20 deletions in the number of occurrences in the ARG
are shown in Table~\ref{tab:rec_dels}.
About half of these deletions are one-base,
occur in non-coding regions, and
are concentrated in the 3' end of the genome.
Four one-base deletions occur
in accessory genes (ORF7a, ORF8, and ORF10).
The two deletions in S1-NTD are known recurrent deletions
associated with antibody escape \cite{McCarthy2021}.

>>>>

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., \cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ''X'', e.g., XA and XBB).

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass QC (STAR Methods).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed a more comprehensive mutational landscape
of SARS-CoV-2 by mapping deletions onto the ARG
using post-hoc parsimony.
We identified important deletion events
(which are inherited by at least 10,000 samples;
Table~\ref{tab:major_dels}).
All of these deletions
(except a one-base deletion at 28,271
above a B.1-labeled node)
are lineage-defining mutations of major VOCs
(Alpha, Delta, Omicron BA.1, BA.2, and BA.4).
Many of these deletions occur in S1-NTD,
a known hotspot of in-frame deletions (REF).
There is a recurrent 1-base deletion at 28,271,
which is in a non-coding region.
<TODO: The 5-base dels and 4-base del are split up.
It should be a 9-base del (11,288–11,296).
I don't see excluded sites among these deleted sites.>

We also found highly recurrent deletions using the ARG.
The top 20 deletions in the number of occurrences
are shown in Table~\ref{tab:rec_dels}.
About half of these deletions affect a single base
and occur in non-coding regions
(concentrated in the 3' end of the genome);
some occur in accessory genes ORF7a and ORF8.
<TODO: Are any of these assembly errors or misalignments?>.
Two of these are known recurrent deletions in S1-NTD,
which are associated with the origins of major VOCs (REF).
